Rhizaria in the oligotrophic ocean exhibit clear temporal and vertical variability.

### 2024-01-29 manuscript/01_intro.qmd
render with line numbers

Rhizaria are an extremely diverse super-group of single celled organisms consisting of several phyla including retaria (foraminifera and radiolaria) and cercozoa. These organisms exist in a wide range of habitats and are widely represented in plankton communities throughout the global ocean. While the taxonomy of these organisms has recently undergone several reclassifications [@biard2022], their presence in ocean ecosystems has been long known to oceanographers. Some of the earliest records of their existence are from oceanographic expeditions in the 19th century [@haekel1887]. Rhizaria are unique members of the plankton and protist community because they can reach large sizes (up to several mm in diameter) and they construct intricate mineral skeletons out of either silica, strontium, or calcium carbonate [@kimoto2015; @suzuki2015; @nakamura2015; @biard2022]. Despite their noticeable morphology and global distribution, Rhizaria were largely understudied throughout the 20th century. The bulk of modern plankton research has focused hard-bodied crustacea which are numerically dominant and easily sampled with nets and preservatives. Fragile organisms like Rhizaria, were difficult to adequately study as they can be destroyed through standard zooplankton sampling techniques. A number of studies in the late 1900s did employ alternative techniques to quantify Rhizaria including diaphragm pumps [@michaels1988] or blue-water SCUBA collections [@caron1984; @caron1995; @bijma1990]. However, the bulk of Rhizaria research was constrained to sediment traps or paleontological studies of sediment [@takahashi1983; @boltovskoy1993]. Only recently has the advent of molecular techniques and in-situ imaging tools ignited a renewed focus on Rhizaria in pelagic ecosystems [@caron2016].

The wave of new data on Rhizaria has facilitated an improved understanding of the significance in ocean ecosystem functions. Firstly, taxonomists have been able to greatly refine the understanding of evolutionary relationships amongst these diverse protists [@aurahs2009; @decelle2012; @decelle2013; @biard2015; @cavalier-smith2018; rev by @biard2022]. DNA metabarcoding studies have revealed insights into the distributional patterns [@decelle2013; @biard2017; @marsbrisbin2020; @blanco-bercial2022 @llopismonferrer2022; @sogawa2022], ecological relationships [@decelle2012; @nakamura2023], and contribution to biogeochemical fluxes [@guidi2016; @gutierrez-rodriguez2019]. Transcriptomic and proteomic approaches also have been used to quantify rhizarian contribution to community metabolism [@cohen2023]. Yet, despite the excellent taxonomic resolution provided by molecular approaches, they do not provide a truly quantitative metric for estimating Rhizarian abundance or biomass. In-situ imaging tools however, offer the ability to observe organisms in the natural state and quantify their abundance [@ohman2019a; @barth-inrev]. While there were early applicaitons of imaging tools to document Rhizaria [@dennett2002], @biard2016's report from a global imaging dataset highlighted the importance of Rhizaria to the total standing stock of marine carbon. Due to their large sizes, ability to concentrate smaller particles (ballasting), and the unique structure of their mineral skeletons, Rhizaria have the potential to massively influence ocean biogeochemical cycling. A number of studies have made large advances in estimating the contribution of Rhizaria to ocean cycling of carbon [@lampitt2009; @stukel2018; @ikenoue2019; @gutierrez-rodriguez2019], silica [@biard2018; @llopismonferrer2021], and strontium [@decelle2013]. Still, Rhizarian ecological roles are not well understood [@biard2022]. This is a major challenge as it is critical to understand the ecological role of plankton to fully incorporate them into biological oceanographic models.

The ecological role of Rhizaria in plankton communities is complicated due to the fact different taxa can exhibit every different trophic modes. As zooplankton, rhizaria are predominately heterotrophic [@biard2022], yet their feeding modes can be quite varied. Phaeodarians (family Cercozoa) are largely thought to be flux-feeders, collecting and feeding on sinking particles [@nakamura2015; @stukel2019]. Alternatively, Retaria can be either exclusively heterotrophic or mixotrophic, utilizing photosynthetic algal symbionts [@anderson2014; @decelle2015]. Mixotrophic foraminifera host a variety of endosymbiont partners [@lee2006; @decelle2015], which are thought to support early and adult life stages and contribute to total primary productivity [@kimoto2015]. Still, foraminifera are omnivorous, possibly even predominately carnivorous with several studies suggesting that they can be effective predators [@anderson1976; @gaskell2019], majoritively consuming live copepods [@caron1984]. Radiolaria have several lineages all which have some taxa who are well known to host symbionts [@biard2022b]. Amongst radiolarians, arguably the most widespread are Collodaria who can be either large solitary cells or form massive colonies, up to several meters in length [@swanberg1981]. All known Collodaria species host dinoflagellate symbionts [@biard2022b] and can contribute substantially to primary productivity, particularly in oligotrophic ocean regions [@caron1995; @dennett2002]. This Collodaria-symbiont association has been suggested as a reason for their high abundances throughout the photic zone of oligotrophic environments globally [@biard2016; @biard2017]. A few Acantharean (Radiolarian order) clades host algal symbionts [@decelle2012; @biard2022b], notably with two clades forming an exclusive relationship with Phaeocystis. However, globally Acantharea are less abundant than Collodaria [@biard2022] and contribute less to total primary productivity [@michaels1995]. This may be due to the fact several clades of Acantharea are cyst-forming and strictly heterotrophs [@decelle2013; @biard2022b]. Furthermore, @marsbrisbin2020 documented apparent predation behavior in Acantharea near the surface, suggesting that there may be a large reliance on carnivory.

Given the high abundances, yet diverse trophic strategies found among Rhizarian taxa, it is reasonable to expect some form of niche partitioning. A number of studies do suggest evidence for vertical zonation between Rhizaria groups according to various trophic strategies. Taxa-specific studies of radiolarians suggest they may be restricted to the euphotic zone [@michaels1988; @boltovskoy2017]. Although some studies report Acantharea in deeper waters [@decelle2013; @gutiérrez-rodríguez2022]. Phaeodarians alternatively, are generally found in the mesopelagic where photosynthesis cannot occur but they can feed on sinking particles [@stukel2018]. In an imaging-based study of the whole Rhizaria community, @biard2020 noted clear patterns in vertical zonation which largely corresponded to different trophic roles. In the oligotrophic ocean, @blanco-bercial2022 also noted that the protist community, including Rhizaria partition along an autotroph and mixotroph to heterotroph gradient with increasing depth in the water column. Yet, few studies have made direct attempts to relate rhizaria abundances to environmental factors [@biard2020]. In part, this is due to the fact few studies have been able to sample Rhizaria in the same location over a consistent timeframe [@michaels1988; @boltovskoy1993; @michaels1995; @hull2011; @gutiérrez-rodríguez2022]. Furthermore, no studies have utilized imaging, arguably the best method for quantifying rhizaria, consistently throughout the full mesopelagic. Given this lack of information, there are many unknowns with respect to Rhizarian ecology, seasonality and phenology across different groups.

>>>>

Rhizaria are an extremely diverse super-group of single-celled organisms consisting of several phyla including Retaria (foraminifera and radiolaria) and Cercozoa. These organisms exist in a wide range of habitats and are widely represented in plankton communities throughout the global ocean. While the taxonomy of these organisms has recently undergone several reclassifications [@biard2022], their presence in ocean ecosystems has been long known to oceanographers. Some of the earliest records of their existence are from oceanographic expeditions in the 19th century [@haekel1887]. Rhizaria are unique members of the plankton and protist community because they can reach large sizes (up to several mm in diameter) and they construct intricate mineral skeletons out of either silica, strontium, or calcium carbonate [@kimoto2015; @suzuki2015; @nakamura2015; @biard2022]. Despite their noticeable morphology and global distribution, Rhizaria were largely understudied throughout the 20th century. The bulk of modern plankton research has focused on hard-bodied crustacea which are numerically dominant and easily sampled with nets and preservatives. Fragile organisms like Rhizaria, were difficult to adequately study as they can be destroyed through standard zooplankton sampling techniques and preserve poorly. A number of studies in the late 1900s did employ alternative techniques to quantify Rhizaria including diaphragm pumps [@michaels1988] or blue-water SCUBA collections [@caron1984; @caron1995; @bijma1990]. However, the bulk of Rhizaria research was constrained to sediment traps or paleontological studies of sediment [@takahashi1983; @boltovskoy1993]. Only recently has the advent of molecular techniques and in-situ imaging tools ignited a renewed focus on Rhizaria in pelagic ecosystems [@caron2016].

The wave of new data on Rhizaria has facilitated an improved understanding of the significance in ocean ecosystem functions. Firstly, taxonomists have been able to greatly refine the understanding of evolutionary relationships amongst these diverse protists [@aurahs2009; @decelle2012; @decelle2013; @biard2015; @cavalier-smith2018; rev by @biard2022]. DNA metabarcoding studies have revealed insights into the distributional patterns [@decelle2013; @biard2017; @marsbrisbin2020; @blanco-bercial2022 @llopismonferrer2022; @sogawa2022], ecological relationships [@decelle2012; @nakamura2023], and contribution to biogeochemical fluxes [@guidi2016; @gutierrez-rodriguez2019]. Transcriptomic and proteomic approaches also have been used to quantify rhizarian contribution to community metabolism [@cohen2023]. Yet, despite the excellent taxonomic resolution provided by molecular approaches, they do not provide a truly quantitative metric for estimating Rhizarian abundance or biomass. In-situ imaging tools however, offer the ability to observe organisms in the natural state and quantify their abundance [@ohman2019a; @barth-inrev]. While there were early applicaitons of imaging tools to document Rhizaria [@dennett2002], @biard2016's report from a global imaging dataset highlighted the importance of Rhizaria to the total standing stock of marine carbon. Due to their large sizes, ability to concentrate smaller particles and the unique structure of their mineral skeletons, Rhizaria have the potential to massively influence ocean biogeochemical cycling. A number of studies have made large advances in estimating the contribution of Rhizaria to ocean cycling of carbon [@lampitt2009; @stukel2018; @ikenoue2019; @gutierrez-rodriguez2019], silica [@biard2018; @llopismonferrer2021], and strontium [@decelle2013]. Still, Rhizarian ecological roles are not well understood [@biard2022]. This is a major challenge as it is critical to understand the ecological role of plankton to fully incorporate them into biological oceanographic models.

The ecological role of Rhizaria in plankton communities is complicated due to the fact different taxa can exhibit every different trophic modes. As zooplankton, rhizaria are predominately heterotrophic [@biard2022], yet their feeding modes can be quite varied. Phaeodarians (family Cercozoa) are largely thought to be flux-feeders, collecting and feeding on sinking particles [@nakamura2015; @stukel2019]. Alternatively, Retaria can be either exclusively heterotrophic or mixotrophic, utilizing photosynthetic algal symbionts [@anderson2014; @decelle2015]. Mixotrophic foraminifera host a variety of endosymbiont partners [@lee2006; @decelle2015], which are thought to support early and adult life stages and contribute to total primary productivity [@kimoto2015]. Still, foraminifera are omnivorous, possibly even predominately carnivorous with several studies suggesting that they can be effective predators [@anderson1976; @gaskell2019], mainly consuming live copepods [@caron1984]. Radiolaria have several lineages all which have some taxa who are well known to host symbionts [@biard2022b]. Amongst radiolarians, arguably the most widespread are Collodaria who can be either large solitary cells or form massive colonies, up to several meters in length [@swanberg1981]. All known Collodaria species host dinoflagellate symbionts [@biard2022b] and can contribute substantially to primary productivity, particularly in oligotrophic ocean regions [@caron1995; @dennett2002]. This Collodaria-symbiont association has been suggested as a reason for their high abundances throughout the photic zone of oligotrophic environments globally [@biard2016; @biard2017]. A few Acantharean (Radiolarian order) clades host algal symbionts [@decelle2012; @biard2022b], notably with two clades forming an exclusive relationship with Phaeocystis. However, globally Acantharea are less abundant than Collodaria [@biard2022] and contribute less to total primary productivity [@michaels1995]. This may be due to the fact several clades of Acantharea are cyst-forming and strictly heterotrophs [@decelle2013; @biard2022b]. Furthermore, @marsbrisbin2020 documented apparent predation behavior in Acantharea near the surface, suggesting that there may be a large reliance on carnivory.

Given the high abundances, yet diverse trophic strategies found among Rhizarian taxa, it is reasonable to expect some form of niche partitioning. A number of studies do suggest evidence for vertical zonation between Rhizaria groups according to various trophic strategies. Taxa-specific studies of radiolarians suggest they may be restricted to the euphotic zone [@michaels1988; @boltovskoy2017]. Although some studies report Acantharea in deeper waters [@decelle2013; @gutiérrez-rodríguez2022]. Phaeodarians alternatively, are generally found in the mesopelagic where photosynthesis cannot occur but they can feed on sinking particles [@stukel2018]. In an imaging-based study of the whole Rhizaria community, @biard2020 noted clear patterns in vertical zonation which largely corresponded to different trophic roles. In the oligotrophic ocean, @blanco-bercial2022 also noted that the protist community, including Rhizaria, partition along an autotroph and mixotroph to heterotroph gradient with increasing depth in the water column. Yet, few studies have made direct attempts to relate rhizaria abundances to environmental factors [@biard2020]. In part, this is due to the fact few studies have been able to sample Rhizaria in the same location over a consistent timeframe [@michaels1988; @boltovskoy1993; @michaels1995; @hull2011; @gutiérrez-rodríguez2022]. Furthermore, no studies have utilized imaging, arguably the best method for quantifying rhizaria, consistently throughout the full mesopelagic. Given this lack of information, there are many unknowns with respect to Rhizarian ecology, seasonality and phenology across different groups.

### 2024-01-29 manuscript/02_methods.qmd
render with line numbers

Data were collected in collaboration with the Bermuda Atlantic Time-Series Study [@michaels1996; @lomas2013] on board the R/V Atlantic Explorer. Cruises were conducted at approximately monthly intervals. Rhizaria individuals were sampled using the Underwater Vision Profiler 5 [UVP5; @picheral2010], a tool which is well established to accurately quantify large Rhizaria [@biard2016; @stukel2018; @stukel2019; @biard2020; @barth2022; @drago2022; @llopismonferrer2022; @panaïotis2023]. The UVP5 was mounted to the sampling rosette and collected data autonomously on routine casts, from which only the downcast data are utilized. The UVP5 was deployed from June-September 2019 then from October 2020 - January 2022. Casts were filtered to only include data collected in the BATS region, far offshore of Bermuda into the Sargasso Sea (Supplemental Figure 1). In general, casts extended to either 200m, 500m, or 1200m deep, with a few extended into the bathypelagic (4500m). However, Rhizaria were only typically found in large abundances throughout the epipelagic and mesopelagic zones. As such, we limit this study to results from the upper 1000m of the water column.

@@

Individual vignettes of Rhizaria images were identified using the classification platform Ecotaxa \[cite\]. Data were pre-sorted according to a . Taxonomic classification were done based on morphology exclusively. While there are sparse taxonomic guides for in-situ images of rhizaria, identification largely relied on descriptions in [@nakamura2015; @suzuki2015; and @biard2020]. Using the aforementioned sources and publicly available ecotaxa projects, I constructed a guide which can be found here: <https://thealexbarth.github.io/media//Project_Items/Oligotrophic_Community/ecotaxa_UVP-guide-stone-lab.pdf>. Broadly, Rhizaria were classified as Foraminifera, Radiolarians (Acantharea or Collodaria), or a variety of Phaeodarian families (Figure 1). When identification could not be confidently made between a few candidate taxa, a less specific label was used. As a result, we have data from "unidentified Rhizaria", which typically were vignettes not distinguishable between Aulacanthidae or Acantharea or "unidentified Phaeodaria", which are clearly phaeodaria but not distinguishable into a family.

![Example images of different Rhizaria taxa. 4mm scale bar shown in lower right. All vignettes are same scale](images/01_taxa.pdf)

The UVP5 samples at \~15Hz rate as it descends the water column and records the exact position of each particle larger than 600$\mu m$. However, identified rhizaria ranged from a 934$\mu m$ Aulacanthidae cell to a Collodarian colony over 10mm in diameter. To confirm that the UVP5 was sampling adequately across all size ranges, an NBSS slope was constructed to identify a drop-off which would indicate poor-sampling at the small size range [@lombard2019; @barth-inrev] . However, it was evident from this analysis that all size ranges were adequately sampled across the size range (Supplemental Figure 2) so no data were excluded. The UVP5 reports the exact depth at which a particle is recorded, however to estimate abundance, observations must be binned over fixed depth intervals. In our deployments had variable descent depths and speeds with more casts descending to 500m than 1000m and descents quicker through the epipelagic than the mesopelagic [see @barth2022 for an extended discussion of UVP5 data processing). For the present study, Rhizarian abundances were estimated in 25m vertical bins, which offer a moderate sampling volume (average 0.948$m^3$ in the epipelagic and 0.589$m^3$ in the mesopelagic) while still maintaining ecologically relevant widths. However, concentrations in a 25m bin would need to be greater than 2.428 indv. $m^{-3}$ and 3.912 indv. $m^{-3}$, in the epipelagic and mesopelagic respectively, to fall below a 10% non-detection risk [@benfield1996; @barth-inrev]. Because we typically observed many rhizaria taxa below these concentrations, we present the 25m binned data to visualize broad-scale average distributions. For quantifying and modelling Rhizaria abundances, we present integrated abundance estimates, with each cast. Due to the variable descent depths of the UVP, data are categorized as epipelagic (0-200m), upper mesopelagic (200-500m), and lower mesopelagic (500-1000m). The average sampling volume integrated through these regions were 7.59$m^2$, 7.06$m^2$, and 11.77$m^2$, with non-detection thresholds at 0.30 indv. $m^{-2}$, 0.33 indv. $m^{-2}$, and 0.20 indv. $m^{-2}$ respectively. All UVP data processing was done using the `EcotaxaTools` package in R (Barth 2023).

@@

Generalized Additive Models (GAMs) were used to assess the relationship between integrated Rhizarian abundance and different environmental factors. GAMs offer the ability to model non-linear and non-monotonic relationships, which can be particularly useful in assessing ecological relationships [@wood2017] and have been successfully applied to Rhizarian ecology [@biard2020]. The `mgcv` package [@wood2001] was used to construct models relating environmental parameters to each taxonomic group's integrated abundance estimates from each cast. To select the most parsimonious model for each analysis, a backwards step-wise approach was taken. First, a full model was fit using any term which may be ecologically relevant. Terms were fit using maximum likelihood with a double penalty approach on unnecessary smooths [@marra2011]. The smoothness parameter was restricted (k = 6) to prevent overfitting the models. At each iteration of the backwards step-wise procedure, the model term with the lowest F score (least statistically significant) was removed. This was repeated until all model terms were statistically significant or the $R^2_{adj}$ was substantially reduced. Models were fit for each region; epipelagic, upper mesopelagic, and lower mesopelagic. In cases where observations were too sparse for a given taxonomic grouping, models were not run.

>>>>

Data were collected in collaboration with the Bermuda Atlantic Time-Series Study [@michaels1996; @lomas2013] on board the R/V Atlantic Explorer. Cruises were conducted at approximately monthly intervals. Rhizaria individuals were sampled using the Underwater Vision Profiler 5 [UVP5; @picheral2010], a tool which is well established to accurately quantify large Rhizaria [@biard2016; @stukel2018; @stukel2019; @biard2020; @barth2022; @drago2022; @llopismonferrer2022; @panaïotis2023]. The UVP5 was mounted to the sampling rosette and collected data autonomously on routine casts, from which only the downcast data are utilized. The UVP5 was deployed from June-September 2019 then from October 2020 - January 2022. Casts were filtered to only include data collected in the BATS region, far offshore of Bermuda into the Sargasso Sea (approximately 31.0$^oN$-32.5$^oN$, 64.25$^oW$-63$^oW$; Supplemental Figure 1). In general, casts extended to either 200m, 500m, or 1200m deep, with a few extended into the bathypelagic (4500m). However, Rhizaria were only typically found in large abundances throughout the epipelagic and mesopelagic zones. As such, we limit this study to results from the upper 1000m of the water column.

@@

Individual vignettes of Rhizaria images were identified using the classification platform Ecotaxa [@picheral]. Data were pre-sorted utilizing a random-forest classifier and pre-trained learning set. Taxonomic classification were done based on morphology exclusively. While there are sparse taxonomic guides for in-situ images of rhizaria, identification largely relied on descriptions in [@nakamura2015; @suzuki2015; and @biard2020]. Using the aforementioned sources and publicly available ecotaxa projects, I constructed a guide which can be found here: [https://thealexbarth.github.io/media//Project_Items/Oligotrophic_Community/ecotaxa_UVP-guide-stone-lab.pdf](#0){.uri}. Broadly, Rhizaria were classified as Foraminifera, Radiolarians (Acantharea or Collodaria), or a variety of Phaeodarian families (Figure 1). When identification could not be confidently made between a few candidate taxa, a less specific label was used. As a result, we have data from "unidentified Rhizaria", which typically were vignettes not distinguishable between Aulacanthidae or Acantharea or "unidentified Phaeodaria", which are clearly phaeodaria but not distinguishable into a family.

The UVP5 samples at \~15Hz rate as it descends the water column and records the exact position of each particle larger than 600$\mu m$. However, identified rhizaria ranged from a 934$\mu m$ Aulacanthidae cell to a Collodarian colony over 10mm in diameter. To confirm that the UVP5 was sampling adequately across all size ranges, an NBSS slope was constructed to identify a drop-off which would indicate poor-sampling at the small size range [@lombard2019; @barth-inrev]. However, it was evident from this analysis that all size ranges were adequately sampled across the size range (Supplemental Figure 2) so no data were excluded. The UVP5 reports the exact depth at which a particle is recorded, however to estimate abundance, observations must be binned over fixed depth intervals. Our deployments had variable descent depths and speeds with more casts descending to 500m than 1000m and descents quicker through the epipelagic than the mesopelagic (see @barth2022 for an extended discussion of UVP5 data processing). For the present study, Rhizarian abundances were estimated in 25m vertical bins, which offer a moderate sampling volume per bin (average 0.948$m^3$ in the epipelagic and 0.589$m^3$ in the mesopelagic) while still maintaining ecologically relevant widths. However, concentrations in a 25m bin would need to be greater than 2.428 indv. $m^{-3}$ and 3.912 indv. $m^{-3}$, in the epipelagic and mesopelagic respectively, to fall below a 10% non-detection risk \[@benfield1996; @barth-inrev\]. Because we typically observed many rhizaria taxa below these concentrations, we present the 25m binned data to visualize broad-scale average distributions. For quantifying and modelling Rhizaria abundances, we present integrated abundance estimates, with each cast. Due to the variable descent depths of the UVP, data are categorized as epipelagic (0-200m), upper mesopelagic (200-500m), and lower mesopelagic (500-1000m). The average sampling volume integrated through these regions were 7.59$m^2$, 7.06$m^2$, and 11.77$m^2$, with non-detection thresholds at 0.30 indv. $m^{-2}$, 0.33 indv. $m^{-2}$, and 0.20 indv. $m^{-2}$ respectively. All UVP data processing was done using the `EcotaxaTools` package in R (Barth 2023).

@@

Generalized Additive Models (GAMs) were used to assess the relationship between integrated Rhizarian abundance and different environmental factors. GAMs offer the ability to model non-linear and non-monotonic relationships, which can be particularly useful in assessing ecological relationships [@wood2017] and have been successfully applied to Rhizarian ecology [@biard2020]. The `mgcv` package [@wood2001] was used to construct models relating environmental parameters to each taxonomic group's integrated abundance estimates from each cast. To select the most parsimonious model for each analysis, a backwards step-wise approach was taken. First, a full model was fit using any term which may be ecologically relevant. Terms were fit using maximum likelihood with a double penalty approach on unnecessary smooths [@marra2011]. The smoothness parameter was restricted (k = 6) to prevent overfitting the models. At each iteration of the backwards step-wise procedure, the model term with the lowest F score (least statistically significant) was removed. This was repeated until all model terms were statistically significant or the $R^2_{adj}$ was substantially reduced. Models were fit for each region; epipelagic, upper mesopelagic, and lower mesopelagic. In cases where observations were too sparse for a given taxonomic grouping, models were not run. All code and full models are available in code, as well as intermediate data products at <https://github.com/TheAlexBarth/RhizariaSeasonality>.
